What We Eval in the Shadows

### 2021-08-11 paper/main.tex
conclusion and discussion

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig,tikz}
\usetikzlibrary{positioning,automata,fit,shapes.geometric,backgrounds,calc}
\usepackage{tabularx}
\input{macro}
\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
\newcommand{\authorcomment}[3]{\xspace\textcolor{#1}{{\bf #2} #3}\xspace}
\newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
\newcommand{\code}[1]{\lstinline |#1|\xspace}
%\renewcommand{\c}[1]{\lstinline |#1|\xspace}
% cf. https://tex.stackexchange.com/a/144640
\makeatletter\let\expandableinput\@@input\makeatother
\lstdefinelanguage{smalleR} {
  morekeywords={
    for,
    if,
    else,
    function
  },
  sensitive=true, % keywords are not case-sensitive
  morecomment=[l]{\#}, % l is for line comment
  morestring=[b]{"} % defines that strings are enclosed in double quotes
}
\lstset{
  language={smalleR},
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
%  framexleftmargin=1mm,
%  framexrightmargin=1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  alsoletter={., _, $},
  breaklines=true
}

@@

\paragraph{Miscellaneous} Many uses of \eval are diverse and do not fit into a
category by themselves. For example, packages for unit testing (testthat,
testit, etc.), code benchmarking (rbenchmark, microbenchmark, etc.), and,
running code in parallel (foreach, doParallel, etc.), use \eval to evaluate
program fragments in a particular way in specific environments. Many numerical
computation and mathematical modeling packages use \eval to evaluate chunks of
code or bodies of functions in specific environments. Consider the
\code{findZeros} function from \code{mosaic} package which is used to compute
the zero of a function.

\begin{figure}[h]
\begin{lstlisting}
findZeros <- function(expr, ..., xlim=c(near-within, near+within), near=0, within=Inf) {
  dots <- list(...)

  pfun <- function(x){  # removed . from name, was .x
    mydots <- dots
    mydots[[rhsVars]] <- x
    eval( lhs(expr), envir=mydots, enclos=parent.frame() )
  }
  searchx <- sort(unique(internal.near   + c(0, leftseq, middleseq, rightseq)))
  y <- sapply( searchx, pfun )
\end{lstlisting}
  \caption{mosaic::findZeros}
\end{figure}

This function can be invoked as: \code{findZeros( A*sin(2*pi*t/P) ~ t,
  xlim=c(0,100), P=50, A=2)}. It will evaluate \code{A*sin(2*pi*t/P)} inside
\code{pfun}. It calls \code{pfun} repeatedly with different values of \code{t}
to find the point at which the expression is zero.

@@

In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of
expressions passed to \eval. Here, we illustrate them in concrete instances.

\begin{compactitem}[---]

  \item \emph{Variable lookup and values, and indexing \c{(X, V, \$)}}
    constitute one of the most common use case for \eval. Let's consider the
    \c{subset} function from the R base library, which allows, among others, to
    select variables in a data frame. For example, given a data frame \c{df}
    with three columns \c{A,B,C}, \c{subset(df, sel=A)} returns a subset of
    \c{df} with just the \c{A} column. The same would happen with \c{subset(df,
      sel=1)}. However, thanks to \eval it can also be also called with
    \c{subset(df, sel=c(1,B,x))} returning a subset of the first column, column
    \c{B} and whatever index the variable \c{x} holds. It is implemented using:
    \lstinline|eval(substitute(sel), nl, parent.frame())| The first argument
    returns the expression passed to \c{sel}, the second is the data frame in
    which this expression will be evaluated, and the last indicates which
    environment should be used to look up bindings not found in \c{nl}.

@@

  \item \emph{Assignments \c{(<-)}} appear in the cases where a developer wants
    to create or update a binding in a specific environment where either the
    name or the value come from some expression. Most cases originate from
    trivial code generation where the assign expression is assembled using
    \c{parse} or \c{substitute}, often in a loop. For example, the
    \c{PopED::get\_rse} uses it to assign a number of variables in the current
    scope: \lstinline|eval(parse(text=paste(default_args[[i]]), "<-", i))|.

@@

  \item \emph{Function definition \c{(FUN)}} is mostly used in conjunction with
    code generation where new functions are synthesized using either \c{parse}
    or \c{substitute}. Some FFI frameworks use this to automatically generate R
    binding to native code. For example, \c{Rcpp}, which bridges R with C++,
    uses \eval to generate R functions for C++ methods:\\
    \lstinline|eval(substitute(function(...) .CppObj$M(...), list(M=...)), env)|.

  \item \emph{Function calls \c{(F(), F(X), F(F(X)))}} represent two very
    frequent patterns: \emph{code generation} and \emph{call reflection}. There
    are numerous cases for code generation.
    %
    One occurring usage (present also in the Kaggle scripts) is to generate
    ad-hoc calls to libraries such as \c{ggplot2} when one needs to
    parameterize data or aesthetics selection from a variable. For example
    \lstinline|ggplot(df, aes(x=A, y=B))| maps X and Y axes in the plot to
    \c{A} and \c{B} columns in \c{df}. It might not be immediately
    clear\footnote{\cf
    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}
    for a ggplot2 user how to modify the call to assign the axis columns from
    variables. Another case is the traditional use of \eval as a mean to reduce
    boilerplate code; simple and repetitive code can easily be replaced with
    judicious use of \eval. For example, the \c{data.table} package uses \eval
    to call the \c{options} function with named arguments taken from a vector
    of strings:

@@

  \item \emph{Block \c{(BLOCK)}} can appear anywhere, but it is almost always
    the case that the block is only passed to \eval rather than being directly
    part of in as in \lstinline|eval({...})|. 
\end{compactitem}

%\todo{Say somewhere that the R community commonly calls call/argument capturing
%(with substitute or match.call), Non-Standard Evaluation.}

\subsection{Discussion}

\Eval in R is used chiefly for meta-programming purposes and to access
environments other than the local one. In Javascript, patterns are ad-hoc, such
as json loading and parsing, and can often be rewritten without \eval, such as a
function or method call, or an assignment to a local variable or an object
property. For R, the data suggests that in some cases, the use of \eval in the
\code{X, V, \$, <-, FUN} and the function calls related to code generation could
replaced by more specific functions. For instance, variable lookup in any
environment can be performed with \code{get}, and assignment in any environment,
with \code{assign}. Building a call and executing can be done with \code{do.call}.
Nevertheless, many of the observed \eval usage in R, particularly, the ones
related to statistical modeling and DSLs are more sophisticated than those
reported for JavaScript.

@@

\subsection{Removing \eval}

Similarly to JavaScript, there are also unnecessary uses of \eval. These uses of
\eval can be replaced with alternate functions yielding the same results. For
example, the \code{PerformanceAnalytics} package contains a function
\code{chart.QQPlot} that uses \eval to resolve a string into function and another
to call it and assign its results into a variable:
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- eval(parse(text=paste("q",d,sep=""))); z <- NULL
  	eval(parse(text=paste("z<-q.f(",dp,",...)")))
  }
\end{lstlisting}
  In both cases, there is no need for \eval and the function can be rewritten as
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- get(paste0("q",d)); z <- q.f(dp, ...)
  }
\end{lstlisting}
  or even to a one-liner \code{do.call(paste0("q",d), c(dp, list(...)))}.

  The \code{copyslots} function from the \code{coin} package uses \eval to copy the
  slots from one \code{S4} object to another. It performs
  \code{eval(str2lang(paste0("target@", s, " <- source@", s)))} for each slot name
  \code{s}. This can be replaced by the expression \code{slot(target, s) <- slot(source,
    s)} which uses the \code{slot<-} and \code{slot} functions to assign and retrieve
  the slots by name.

  Another examples is in the \code{configr::config.funs.par} function whose
  simplified definition is shown below.

\begin{lstlisting}
config.funs.par <- function(fun = "", ...) {
  args.all <- as.list(match.call()); args.all <- args.all[names(args.all) != ""];
  parameters <- list()
  text.1 <- sprintf("parameters <- names(as.list(args(%s)))", fun)
  text.2 <- "parameters <- parameters[parameters != '']"
  text.3 <- "parameters <- args.all[names(args.all) %in% parameters]"
  for (j in c(text.1, text.2, text.3)) {
    eval(parse(text = j))
  }
  return(parameters)
}
\end{lstlisting}

  The function appears to use \eval since the function name \code{fun} is received
  as a string. This string can be resolved to the function object using the
  \code{get} function and the contents of \code{text.1}, \code{text.2}, and \code{text.3}
  can be written directly without the need to convert them to text for
  evaluation using \eval.

The problem is that these cases are not easy to spot. Because of the
limits of dynamic analysis, we only have a partial coverage and cannot determine
if for example, \eval sites classified as \c{X, V} will not be eventually called
with an expression in a \c{F(), F(X)} or \c{F(F(X))} category.

\subsection{Alternatives to \eval}

An alternative implementation of \eval, \c{eval\_tidy}, is provided by the
\c{rlang} package. This forms the base of the tidy evaluation framework used by
the libraries collectively referred to as the \c{tidyverse}~\cite{tidyverse}. It
differs from the \c{eval} function in two ways. First, it supports evaluation of
Quosures, which are also custom objects used by the \c{tidyverse} libraries for
metaprogramming. Quosures package expression with an evaluation environment. The
\c{tidy\_eval} function extracts this expression and evaluates it in that
environment, ignoring the environment passed to it explicitly as an argument.
Second, the \c{eval\_tidy} function also accepts a data mask argument.This is a
set of bindings supplied by the user as a data frame or a named vector or list.
The data mask environment take precedence over the environment passed for
evaluation. Since the expression is always evaluated in the data mask, it has
different semantics compared to \c{base::eval}. Effects such as assignment using
\c{<-} happen in the data mask environment. Expressions such as \c{return()} do
not work since the data mask environment does not correspond to a frame on the
call stack.

@@

In this paper, we provide a large-scale study of the use of the \eval function
in the R programming language. It is based primarily on a corpus of
\CranRunnableScripts scripts extracted from \CranPackages R packages. As
expected, \eval is used mainly in libraries. We have found just a few \eval
call sites in scripts coming from a large corpus of Kaggle representing code
written by R practitioners. We used an instrumented version of the R virtual
machine, \rdyntrace, to capture all \eval calls, including the relevant
side effects that happen within these calls. Several insights can be learned
from the data.
%
First, \eval is widely used in the R core packages that come bundled with the
language. Because of that, there is hardly any R code that would not use it.
\eval is used to implement some basic functionality such as package loading.
Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio of
the available \CranPackages packages.
%
The code executed by R \eval has unusual expressive power. It can essentially
modify anything anywhere, which is wary for both R developers and R compiler.
The shapes of expressions passed to \eval are extremely diverse, ranging from
simple variable lookups to complex blocks or the creation of new functions.
Luckily, most sites only see a single shape. \eval unlashes its power and shows
up its singularity compared to \eval in other languages with the diversity of
environments in which it can evaluate its expressions. While most of the time,
the \eval environment is the local environment, it is also often the global
environment, a synthetic environment, or even a package environment. Luckily, as
for the expressions passed to \eval, the environment passed to \eval for one
given site is easily predictable with the first call to the site.

The expressive power of \eval shines even more so in the case of side effects
that can quite literally happen anywhere. The data, however, suggest that this
is not the case. The majority of eval sites do not do any observable side
effects, and when they do, it happens in a predictable environment.

The origins of the expression passed to \eval, either constructed from language
expressions, or stemming from reflection, or built by parsing strings.
Reflection and constructed expressions dominate, which suggests the primary use
of \eval is meta-programming.

Similarly to \citep{ecoop11}, we started looking into \evals in R with the
ambition to replace them with other features that are more amenable to static
analysis. The data, unfortunately, suggests that replacing \eval is going to be
a lengthy journey. Even if we disregard core packages as they are part of the
language and the \eval use is stable, in CRAN \eval is used widely and in more
sophisticated ways than was the case in JavaScript. There the vast majority of
sites fitted an extremely simple categorization based on simple regular
expressions. Not so in R. While some usage, especially the one which calls \eval
with an expression parsed from string, can be categorized and often replaced by
equivalent yet more disciplined and safer code, the case of \eval coming from
reflection is much harder.

Even though we reckon that \eval has more expressive power than macros, as it
can access runtime information such as its call stack, adding a macro system to
R to replace \eval for all meta-programming purposes would help static analysis
as macros can be expanded at analysis-time.

>>>>

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig,tikz}
\usetikzlibrary{positioning,automata,fit,shapes.geometric,backgrounds,calc}
\usepackage{tabularx}
\input{macro}
\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
\newcommand{\authorcomment}[3]{\xspace\textcolor{#1}{{\bf #2} #3}\xspace}
\newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
\newcommand{\code}[1]{\lstinline |#1|\xspace}
\renewcommand{\k}[1]{\lstinline |#1|\xspace}
% cf. https://tex.stackexchange.com/a/144640
\makeatletter\let\expandableinput\@@input\makeatother
\lstdefinelanguage{smalleR} {
  morekeywords={
    for,
    if,
    else,
    function
  },
  sensitive=true, % keywords are not case-sensitive
  morecomment=[l]{\#}, % l is for line comment
  morestring=[b]{"} % defines that strings are enclosed in double quotes
}
\lstset{
  language={smalleR},
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
%  framexleftmargin=1mm,
%  framexrightmargin=1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  alsoletter={., _, $},
  breaklines=true
}

@@

@@

\emph{Variable lookup and values, and indexing \c{(X, V, \$)}}
    constitute one of the most common use case for \eval. Let's consider the
    \c{subset} function from the R base library, which allows, among others, to
    select variables in a data frame. For example, given a data frame \c{df}
    with three columns \c{A,B,C}, \c{subset(df, sel=A)} returns a subset of
    \c{df} with just the \c{A} column. The same would happen with \c{subset(df,
      sel=1)}. However, thanks to \eval it can also be also called with
    \c{subset(df, sel=c(1,B,x))} returning a subset of the first column, column
    \c{B} and whatever index the variable \c{x} holds. It is implemented using:
    \lstinline|eval(substitute(sel), nl, parent.frame())| The first argument
    returns the expression passed to \c{sel}, the second is the data frame in
    which this expression will be evaluated, and the last indicates which
    environment should be used to look up bindings not found in \c{nl}.

@@

 \emph{Assignments \c{(<-)}} appear in the cases where a developer wants
    to create or update a binding in a specific environment where either the
    name or the value come from some expression. Most cases originate from
    trivial code generation where the assign expression is assembled using
    \c{parse} or \c{substitute}, often in a loop. For example, the
    \c{PopED::get\_rse} uses it to assign a number of variables in the current
    scope: \lstinline|eval(parse(text=paste(default_args[[i]]), "<-", i))|.

@@

 \emph{Function calls \c{(F(), F(X), F(F(X)))}} represent two very
    frequent patterns: \emph{code generation} and \emph{call reflection}. There
    are numerous cases for code generation.
    %
    One occurring usage (present also in the Kaggle scripts) is to generate
    ad-hoc calls to libraries such as \c{ggplot2} when one needs to
    parameterize data or aesthetics selection from a variable. For example
    \lstinline|ggplot(df, aes(x=A, y=B))| maps X and Y axes in the plot to
    \c{A} and \c{B} columns in \c{df}. It might not be immediately
    clear\footnote{\cf
    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}
    for a ggplot2 user how to modify the call to assign the axis columns from
    variables. Another case is the traditional use of \eval as a mean to reduce
    boilerplate code; simple and repetitive code can easily be replaced with
    judicious use of \eval. For example, the \c{data.table} package uses \eval
    to call the \c{options} function with named arguments taken from a vector
    of strings:

@@

@@

In R, \eval is used chiefly for meta-programming and access remote environments.
Unlike, JavaScript which had a small set of well-defined patterns that could be
rewritten without \eval, our results suggest that \eval is an integral part of
programming in R.

The cases where \eval appears to be overkill include simple expression shapes
such as \c X, \c V, \$, \c{<-}, and \c{FUN}. Variable lookup in a remote
environment can be performed by the builtin \code{get} function, a similar case
can be made for assignment and the \code{assign} function. There is a dedicated
\c{do.call} function to build function calls. Unnecessary uses of \eval
often can be uncovered by manual analysis.

\begin{lstlisting}
PerformanceAnalytics::chart.QQPlot <- function (d="norm",dp,...) {
   q.f <- eval(parse(text=paste("q",d)))
   z <- NULL
   eval(parse(text=paste("z<-q.f(",dp,",...)")))
  }
\end{lstlisting}

The above uses \eval to first lookup a target function name from a constructed
string, the it constructs a call to the target that include arguments to the
current functions. \Eval is overkill for this code, a semantically equivalent
function body uses \c{get} to perform the lookup and then calls directly.

\begin{lstlisting}
 function (R, d="norm", dp, ...) {
   q.f <- get(paste0("q",d))
   z <- q.f(dp, ...)
 }
\end{lstlisting}

Another example, from \c{coin::copyslots}, constructs an expression that access
fields of an object and copy them. This can be replaced by symbolic field reads
and writes.

\begin{lstlisting}
  eval(str2lang(paste0("target@", s, " <- source@", s)))
\end{lstlisting}

As a last example, \code{configr::config.funs.par} construct three expressions
that are then fed to \eval. Semantically equivalent code would use \c{get}
to lookup \c{fun}, and then directly execute the three expressions.

\begin{lstlisting}
 config.funs.par <- function(fun="",...) {
   args.all <- as.list(match.call())
   args.all <- args.all[names(args.all) != ""]
   parameters <- list()
   t1 <- sprintf("parameters <- names(as.list(args(%s)))", fun)
   t2 <- "parameters <- parameters[parameters != '']"
   t3 <- "parameters <- args.all[names(args.all) %in% parameters]"
   for (j in c(t1,t2,t3))
    eval(parse(text = j))
   return(parameters)
 }
\end{lstlisting}

The challenge for automation is that such examples are not easy to find. Dynamic
analysis is limited due to coverage issues, and static analysis of R remains an
open problem.

We have not yet touched on alternative implementations of \eval. The tidyverse
libraries include their own implementation calls
\c{rlang::eval\_tidy}~\cite{tidyverse}. This departs from \eval in two ways.
First, it supports evaluation of quosures, custom objects used by for
metaprogramming that bundle expressions with an environment. \c{eval\_tidy}
accepts a data mask argument, a set of bindings such as a dataframe or list,
which takes precedence over the environment. Effects such as assignments happen
in the data mask. Expressions such as \c{return()} do not work since the data
mask does not correspond to a frame on the call stack.

@@

The \eval functions used widely and in varied ways in R. The function is an
essential tool for language implementers. Our analysis observed that the base
libraries heavily depend on \eval and any R program will end up calling it
through the core of the language. Independently developed libraries often use
\eval in subtle ways, mostly to perform macro-programming tasks. Our review of
code written by less experienced users suggest that \eval is exceedingly rarely
used. So, it is fair to conclude that majority of R programmers never have to
write a call to \eval, but that the code they write would not run without it.

\Eval is thus a challenge for automated program understanding. Program analysis
and transformation tools or compilers must assume the worst when faced with code
that calls this function. We have observed all sorts of side-effects with
various degree of visibility.

While our results are not encouraging in general, we have observed many cases
where \eval is used in a disciplined and predictable ways. While in the general
case \eval is hell, there are many cases where it is just a function. It does
not appear that a general purpose replacement for \eval is possible, in future
work we hope to focus on special cases and propose tools that target particular
subsets of call sites with some common properties.
